A general Nullstellensatz for generalized spaces

### 2019-03-11 paper-qcoh.tex
Proofs of the main theorems

To state what it means for a~$\TT$-structure in a topos~$\E$ to verify the
axioms of~$\TT$, rendering it a model, the \emph{internal language} of~$\E$ is
used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
\models \alpha$'' to mean that the statement~$\alpha$ holds from the internal
point of view of~$\E$. Since this language is a form of a higher-order
dependently-typed intuitionistic logic, the classifying topos~$\Set[\TT]$ can
be regarded as a higher-order completion of the geometric theory~$\TT$. The
generic model enjoys the universal property that any~$\TT$-model in any
(Grothendieck) topos~$\E$ is the pullback of~$U_\TT$ along an essentially
unique geometric morphism~$\E \to \Set[\TT]$.
\medskip

@@

We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
follows: \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
algebraic certificate we now have a logical certificate, a \emph{proof}
of~$\sigma$. However, this generalized statement is typically false, even for
the generic model~$U_\TT$: The statement
\begin{multline*}
  \qquad\qquad\qquad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
    \text{if~$\sigma$ holds for~$U_\TT$ then~$\ul{\TT}$
    proves~$\sigma$}\urcorner\qquad\qquad\qquad
\end{multline*}
does not hold.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
by~$\TT$, obtained by pulling back the set of sorts, the set of function
symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. For
instance, if~$\TT$ is the theory of rings, then from the internal point of view
of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
More details will be given in Section~\ref{sect:review}. The corner quotes indicate
that for sake of readability, the translation into formal language is to be
carried out by the reader. \par The displayed statement is much stronger than
the statement that for any geometric sequent~$\sigma$, if $\Set[\TT] \models
\speak{$\sigma$ holds for~$U_\TT$}$ then~$\TT$ proves~$\sigma$. This latter statement,
where the universal quantifier and the ``if \ldots then'' have been pulled out,
is true.}

For instance, let~$\TT$ be the theory of rings and let~$\sigma$ be
the sequent~$(\top \vdash 1 + 1 = 0)$. Since there is an intuitionistic proof
that~$\TT$ does not prove~$\sigma$ and toposes are sound with respect to
intuitionistic logic, the statement~$\speak{\ul{\TT} proves~$\sigma$}$ is false
from the internal point of view of~$\Set[\TT]$. However, it is not the case
that the statement~$\speak{$1 + 1 = 0$ in~$U_\TT$}$ is false from the internal
point of view. In fact, this statement holds in a nontrivial slice
of~$\Set[\TT]$, the open subtopos coinciding with the classifying topos of
the theory of rings of characteristic two.

@@

\begin{proof}It is obvious that~(1) implies~(2) implies~(3).

@@

@@

@@

@@

@@

>>>>

To state what it means for a~$\TT$-structure in a topos~$\E$ to verify the
axioms of~$\TT$, rendering it a model, the \emph{internal language} of~$\E$ is
used, roughly reviewed in Section~\ref{sect:review} below. We write~``$\E
\models \alpha$'' to mean that the formula~$\alpha$ holds from the internal
point of view of~$\E$. Since this language is a form of a higher-order
dependently-typed intuitionistic logic, the classifying topos~$\Set[\TT]$ can
be regarded as a higher-order completion of the geometric theory~$\TT$. The
generic model enjoys the universal property that any~$\TT$-model in any
(Grothendieck) topos~$\E$ is the pullback of~$U_\TT$ along an essentially
unique geometric morphism~$\E \to \Set[\TT]$.
\medskip

@@

We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
follows: \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
algebraic certificate we now have a logical certificate, a \emph{proof}
of~$\sigma$. However, this generalized statement is typically false, even for
the generic model~$U_\TT$: The statement
\begin{multline*}
  \qquad\qquad\qquad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
    \text{if~$\sigma$ holds for~$U_\TT$ then~$\ul{\TT}$
    proves~$\sigma$}\urcorner\qquad\qquad\qquad
\end{multline*}
does not hold.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
by~$\TT$, obtained by pulling back the set of sorts, the set of function
symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. For
instance, if~$\TT$ is the theory of rings, then from the internal point of view
of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
More details will be given in Section~\ref{sect:review}. The corner quotes indicate
that for sake of readability, the translation into formal language is to be
carried out by the reader. \par The displayed statement is much stronger than
the statement that for any geometric sequent~$\sigma$, if $\Set[\TT] \models
\speak{$\sigma$ holds for~$U_\TT$}$ then~$\TT$ proves~$\sigma$. This latter statement,
where the universal quantifier and the ``if \ldots then'' have been pulled out,
is true.} In this sense~$\Set[\TT]$ does not believe that~$U_\TT$ is the
generic~$\ul{\TT}$-model.

A concrete counterexample is as follows. Let~$\TT$ be the theory of rings and let~$\sigma$ be
the sequent~$(\top \vdash 1 + 1 = 0)$. Since there is an intuitionistic proof
that~$\TT$ does not prove~$\sigma$ and toposes are sound with respect to
intuitionistic logic, the statement~$\speak{\ul{\TT} proves~$\sigma$}$ is false
from the internal point of view of~$\Set[\TT]$. However, it is not the case
that the statement~$\speak{$1 + 1 = 0$ in~$U_\TT$}$ is false from the internal
point of view. In fact, this statement holds in a nontrivial slice
of~$\Set[\TT]$, the open subtopos coinciding with the classifying topos of
the theory of rings of characteristic two.

@@

\begin{proof}It is trivial that~(1) implies~(2) implies~(3).

@@

\begin{lemma}\label{lemma:truth-to-provability}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a geometric formula over the
signature of~$\TT$ in a context~$\vec x$. Then
\[ \{ \vec x\_ \alpha \} \models \speak{$\ul{\TT}/U_\TT$ proves $(\top
\vdash_{[]} \alpha)$}, \]
where the free variables~$\vec x$ occurring in~$\alpha$ are interpreted by
their generic values over~$\{ \vec x\_ \alpha \}$.
\end{lemma}

% XXX fill in proof

\begin{lemma}\label{lemma:locally-constant}
Let~$\TT$ be a geometric theory.
Let~$\varphi$ be a section of the sheaf of geometric$^\star$ formulas over the signature
of~$\ul{\TT}/U_\TT$ over a stage~$A$. Then there is a covering~$(A_i \to A)_i$
of~$A$ such that for each index~$i$, there is a formula~$\varphi_i$ over the signature
of~$\TT/U_\TT(A_i)$ such that~$A_i \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi_i)$}$.
\end{lemma}

\begin{proof}By passing to a covering, we may suppose that~$\varphi$ is given
by an (external) geometric formula over the signature
of~$\ul{\TT}(A)/U_\TT(A)$. Any function symbol and relation symbol
of~$\ul{\TT}(A)$ occuring in~$\varphi$ is locally given by a symbol in~$\TT$.
Hence the claim would be trivial if~$\varphi$ were a coherent formula, for in
this case we would just have to pass to further coverings, one for each
occuring symbol, a finite number of times in total.

However, in general, we cannot conclude as easily. Write~$A = \{ \vec x\_
\alpha \}$. Let~$R$ be a relation symbol of~$\ul{\TT}(A)$ occuring
in~$\varphi$. Then there is a covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$ such that, restricted to~$\{
\vec y_j\_ \alpha_j \}$, $R$ is given by a symbol~$R_j$ of~$\TT$. To construct
the desired formula~$\varphi'$, we replace any such occurrence~$R(\ldots)$
in~$\varphi$ by
\[ \bigvee_j \bigl((\exists \vec y_j\_ \theta_j) \wedge R_j(\ldots)\bigr). \]
In a similar vein we treat any occurrence of function symbols.

The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_j\_
\theta_j) \seq{\vec x} \bigvee_j \{ \top \,|\, \text{$(\exists \vec y_j\_ \theta_j)$
holds for~$U_\TT$} \}\bigr)$} \]
which in turn follows from Lemma~\ref{lemma:truth-to-provability} exploiting
that~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec y_k\_ \theta_k)
\seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
\end{proof}

\begin{thm}\label{thm:nullstellensatz}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
geometric$^\star$ sequent~$\sigma$ over the signature of~$\ul{\TT}/U_\TT$, the
following statements are equivalent:
\begin{enumerate}
\item The sequent~$\sigma$ holds for~$U_\TT$.
\item The sequent~$\sigma$ is provable$^\star$ modulo~$\ul{\TT}/U_\TT$.
\end{enumerate}
\end{thm}

\begin{proof}The direction~$(2) \Rightarrow (1)$ is immediate because~$U_\TT$ is, from the
internal point of view of~$\Set[\TT]$, a~$\ul{\TT}/U_\TT$-model. Hence even the
following stronger statement holds internally: \emph{For any geometric
sequent~$\sigma$, if~$\ul{\TT}/U_\TT$ proves~$\sigma$, then~$\sigma$ holds
for~$U_\TT$.}

For the direction~$(1) \Rightarrow (2)$ we have to verify that, given any
stage~$A$ and any section~$\sigma$ of the sheaf of
geometric$^\star$ sequents over~$A$, if~$A \models
\speak{$\sigma$ holds for~$U_\TT$}$ then~$A \models
\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$}$. By
Lemma~\ref{lemma:locally-constant} we may suppose that~$\sigma$ is an
(external) geometric sequent over the signature of~$\TT/U_\TT(A)$.

Writing~$A = \{ \vec x\_ \alpha \}$ and~$\sigma = (\varphi \seq{\vec y} \psi)$,
we have~$\{ \vec x\_ \alpha \} \models \forall y\_ (\varphi \Rightarrow \psi)$,
hence~$\{ \vec x, \vec y\_ \alpha \wedge \varphi \} \models \psi$. Thus~$\TT$
proves~$(\alpha \wedge \varphi \seq{\vec x, \vec y} \psi)$. This proof can be
pulled back from~$\Set$ to~$\Set[\TT]/A$ to show~$A \models
\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\alpha
\wedge \varphi \seq{\vec x, \vec y} \psi)$}$. By
Lemma~\ref{lemma:truth-to-provability}, we also have~$A \models
\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\top \seq{[]} \alpha)$}$ (where the free
variables occuring in~$\alpha$ are interpreted as the generic values available
over~$A$), hence~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi
\seq{\vec y} \psi)$}$.
\end{proof}

% XXX: check contexts.

@@

\begin{thm}\label{thm:characterization}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom
scheme~\eqref{eq:nullstellensatz}.
\end{enumerate}
\end{thm}

\begin{proof}The direction~$(1) \Rightarrow (2)$ is immediate, as the internal
language of~$\Set[\TT]$ validates first-order intuitionistic logic and the
axiom scheme~\eqref{eq:nullstellensatz} by Theorem~\ref{thm:nullstellensatz}.

For the converse direction, we note that, given a geometric sequent~$\sigma$
over the signature of~$\TT$, the statement~$\speak{$\ul{\TT}/U_\TT$
proves$^\star$~$\sigma$}$ can be expressed as a geometric formula over the
signature of~$\TT$. Applying this observation successively to subformulas
of the given formula~$\alpha$, Theorem~\ref{thm:nullstellensatz} on the
semantic side and the axiom scheme~\eqref{eq:nullstellensatz} imply that we may
assume that~$\alpha$ is in fact a geometric formula. Hence we are reduced to
the basic fact that, for geometric formulas~$\varphi$, $\Set[\TT] \models
\varphi$ implies that~$\TT$ proves~$\varphi$.
\end{proof}

% XXX coherent case

@@

\section{The generalization to the higher-order case}

\begin{thm}\label{thm:higher-order-nullstellensatz}
Let~$\TT$ be a geometric theory. Let~$x_1\?X_1,\ldots,x_n\?X_n$ be a context over the
signature of~$\TT$. Then the canonical morphism
\[ \mathrm{Form}^\star_{\vec x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
\cdots \times X_n) \]
sending the equivalence class of a geometric$^\star$ formula~$\varphi$ over the
signature of~$\ul{\TT}/U_\TT$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi
\}$ is an isomorphism.
% XXX explain equivalence relation
\end{thm}

An immediate corollary of Theorem~\ref{thm:higher-order-nullstellensatz} is the
Nullstellensatz of Theorem~\ref{thm:nullstellensatz}. Indeed, arguing
internally to~$\Set[\TT]$, let~$\sigma = (\varphi \seq{\vec x} \psi)$ be a
geometric$^\star$ sequent over the signature of~$\ul{\TT}/U_\TT$. Assume
that~$\sigma$ holds for~$U_\TT$. Then the subsets~$\{ (\vec x) \,|\, \varphi
\}$ and~$\{ (\vec x) \,|\, \varphi \wedge \psi \}$ are equal. Hence, by
Theorem~\ref{thm:higher-order-nullstellensatz}, the formulas~$\varphi$
and~$\varphi \wedge \psi$ are provably~$^\star$ equivalent.
Thus~$\ul{\TT}/U_\TT$ proves$^\star$~$(\varphi \seq{\vec x} \psi)$.

% XXX corollary: function spaces

% XXX prove higher-order characterization?

@@

% Zariski

% Kock--Lawvere in SDG

@@

* Discuss higher-order case
